---
title: "Class 12: Population Scale Analysis Homework"
author: "Mankeerat Rataul"
format: pdf
---



> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
dataset <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")

table(dataset$geno)
nrow(dataset)
```
The above shows the sample size for each genotype, the total dataset consisting of 462 datapoints. 

```{r}
library(tidyverse)
AGonly <-  dataset %>%
  arrange(dataset$geno) %>%
  filter(geno == "A/G")

AAonly <-  dataset %>%
  arrange(dataset$geno) %>%
  filter(geno == "A/A")

GGonly <-  dataset %>%
  arrange(dataset$geno) %>%
  filter(geno == "G/G")

summary(AGonly)

summary(AAonly)

summary(GGonly)
```
For A/G individuals the median expression for ORMDL3 is 25.065. For A/A, it is 31.25, and for G/G, it is 20.074. 


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(dataset, aes(x=geno, y=exp)) +
  geom_boxplot() +
  labs(y="ORMDL3 expression", x="Genotype", title="ORMDL3 expression based on SNP")
```
This SNP seems to affect the expression of ORMDL3, as G/G has much lower expression compared to A/A it seems, with A/G bridging the middle. Visually this looks like there is a decrease due to the G allele.