---
title: "Class 5: ggplot2 viz"
author: "Mankeerat Rataul"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like `plot()` and **ggplot2** which is an add on package. 

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head(cars)
plot(cars)

```
Before I can use ggplot2 I need to install it on my computer.

To install it, we can use the function `install.packages()`, however we never run this function in the document because it would redownload it constantly. Instead this is done in the console.

Below are the 3 commands used for the libraries:

```{r}

# install.packages("ggplot2")
# install.packages("gapminder")
# install.packages("dplyr")

```
Once installed, we need to load up the package into our R brain:

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(gapminder)
library(dplyr)
```


The main function in ggplot2's package is called `ggplot()`

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

- the **data** (data.frame of what we want to plot)
- the **aes** (how data maps onto plot)
- the **geom** (how we want the plot to be drawn)

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

## Add some customization/features

Let's add a trendline that shows the relationship between speed + distance. 

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  theme_light() +
  labs(title="Stopping Distance of Old Cars", x="Speed (MPH)", y="Distance(ft)")
```

## Gene Expression Figure

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)



head(genes[genes$State == "up",])

table(genes$State)


round( table(genes$State)/nrow(genes) * 100, 2 )
```

```{r}
colnames(genes)
```


```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() +
  theme_light() +
  labs(x="No drug", y="Drug", title="Expression chnages upon GLP-1 inhibitor")

p + scale_color_manual( values=c("blue","gray","orange"))
```

## Going further into the GapMinder dataset

Here is the read of it

```{r}
head(gapminder)
```

>Q. How many entries are there?

```{r}
nrow(gapminder)
```
>Q. How many countries are in this dataset?

```{r}
length(unique(gapminder$country))
```

Let's make our first plot of this gapminder dataset

```{r}
p <- ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point(alpha=0.3) + 
  theme_light() + 
  labs(x="GDP Per Capita", y="Life Expectancy (years)", title="Life expectancy in relation to GDP Per Capita")
p
```

I can add more layers to change the call of p

```{r}
p +
  facet_wrap(~gapminder$continent)
```

Make a plot for 1977 and 2007
```{r}
gapminder_1977 <- p +
  gapminder %>% filter(year==1977) +
  labs(title="Life expectancy in relation to GDP Per Capita, 1997")

gapminder_2007 <- p + 
  gapminder %>% filter(year==2007) + 
  labs(title="Life expectancy in relation to GDP Per Capita, 2007")

gapminder_1977
gapminder_2007
```
>Q. Make a histogram of lifeExp colored by continent (try using `fill-continent` or `col=continent`)

```{r}
ggplot(gapminder) + 
  aes(lifeExp, fill=continent) +
  geom_histogram()

```

